---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r}
# load data file
mydata <- read.csv("activity.csv")
# process data to convert date column to date class
mydata$date <- as.Date(mydata$date)
```




## What is mean total number of steps taken per day?


Make a histogram of the total number of steps taken each day


```{r}
# make a new dataframe without NA values
cleanData <- mydata[!is.na(mydata$steps), ]
# load plyr library
library(plyr)
# group values by date
groupedByDate <- ddply(cleanData, ~date, summarise, sum = sum(steps))
# construct the histogram
hist(groupedByDate$sum, xlab = "Total Steps", main = "Histogram total steps per day", col = "darkmagenta")
```


Calculate and report the mean and median total number of steps taken per day


```{r}
# Calculate mean value
mean(groupedByDate$sum)

# Calculate median value
median(groupedByDate$sum)
```


The **mean** of the distribution **is: 10766.19**
and its **median is: 10765**




## What is the average daily activity pattern?


Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r}
groupedByInterval <- ddply(cleanData, ~interval, summarise, mean = mean(steps))
with(groupedByInterval, plot(interval, mean, type = "l", ylab = "Average number of steps", xlab = "Interval", main = "Average daily activity", col = "darkred"))
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
# first find max value
maxVal <- max(groupedByInterval$mean)
print(maxVal)

# locate the line containing this value
maxLine <- groupedByInterval[groupedByInterval$mean == maxVal, ]

# find the interval
maxInterval <- maxLine$interval
print(maxInterval)
```


**The maximum number of steps (on average across all the days) is maxVal: 206.1698 **
**and it is contained in interval maxInterval: 835**




## Imputing missing values


Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)


```{r}
# calculate sum of missing values
sum(is.na(mydata$steps))
```


Since there are missing values in the dataset, they need to be filled and I do that by substituting the mean


```{r}
# create a new dataset
newdata <- mydata
# locate tha NAs
missingsteps <- is.na(newdata$steps)
# convert interval(s) to factor(s)
newdata$interval <- factor(newdata$interval)
groupedByInterval$interval <- factor(groupedByInterval$interval)

# fill newdata on missing values wherever we have a missing value, fill it
# from groupedByInterval$mean column (steps) where newdata$interval ==
# groupedByInterval$interval (intervals are factors)
newdata[missingsteps, "steps"] <- groupedByInterval[newdata[missingsteps, "interval"], "mean"]
```


Reconstruct the histogram and compare your results


```{r}
# group values by date
groupedByDate2 <- ddply(newdata, ~date, summarise, sum = sum(steps))

# construct the histogram
hist(groupedByDate2$sum, xlab = "Total Steps", main = "Histogram total steps per day", col = "darkgreen")

# Calculate mean value
mean(groupedByDate2$sum)

# Calculate median value
median(groupedByDate2$sum)
```




**We notice that mean stays the same and median increases by one**
**However as we see from the histogram the impact is very 'light'**





## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
# add a new column containing day of week
newdata$weekday = weekdays(newdata$date)
# add a new column containing either Weekday OR Weekend
newdata$weekday.type <- ifelse(newdata$weekday == "Saturday" | newdata$weekday == "Sunday", "Weekend", "Weekday")
# convert column to factor
newdata$weekday.type <- factor(newdata$weekday.type)
```

Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
# make a new dataset grouping data by interval and weekday.type
groupedBy.Interval.WeekDay <- ddply(newdata, ~interval + weekday.type, summarise, mean = mean(steps))

# For a nice plot 'unfactor' interval. Be careful to convert to characters
# first, or else you get the level values 1,2,3...
groupedBy.Interval.WeekDay$interval <- as.numeric(as.character(groupedBy.Interval.WeekDay$interval))
library(lattice)
xyplot(mean ~ interval | weekday.type, groupedBy.Interval.WeekDay, type = "l", layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")

```


**We notice that mean stays the same and median increases by one in the above comparison.**
**However as we see from the time series the impact is very 'light'**
